Homoiconicity Without Template Haskell

### 2016-06-09 paper/paper.tex
more paper edits

\usepackage{amsmath,amsfonts}

@@

\usepackage{courier}
\usepackage{listings}
\lstset{
    language=Haskell,
    basicstyle=\ttfamily\footnotesize,
    keywordstyle=\ttfamily\footnotesize,
    %basicstyle=\footnotesize,
    literate={->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
             {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
    %keywordstyle=\lst@ifdisplaystyle\color{blue}\fi,
    %commentstyle=\color{gray}
}
\lstMakeShortInline|

@@

@@

@@

@@

Imagine you're writing a machine learning library in Haskell.
%You decide to start by implementing logistic regression.
%The first step is to implement the log-logistic function
%Logistic regression is one of the most important machine learning algorithms,
%and to implement it, you'll need to implement the log-logistic function
%To implement the classic logistic regression algorithm,
%you begin by implementing the log-logistic function
At some point, you'll need to implement the log-logistic function
\begin{equation}
f(x) = \log\left(\frac{1}{1+\exp{(-x)}}\right)
\label{eq:log-logistic}
\end{equation}
This function is used in logistic regression and many families of neural networks.
As a first attempt, you might write:
\begin{lstlisting}
logLogistic1 :: Floating x => x -> x
logLogistic1 x = log(1/(1+exp(-x)))
\end{lstlisting}
Unfortunately, this straightforward implementation is numerically unstable.
When specialized to the |Double| type,
the |exp| function overflows to infinity for inputs greater than (approximately) $709$.
%This overflow causes |logLogistic| to take the |log| of 0, which is undefined.
Therefore, on an input of $-710$,
|logLogisitic| will take the |log| of 0,
returning negative infinity when the true answer is actually very close to $-710$.
Such an inaccurate answer is unacceptable,
so the function must be rewritten.

@@

In this section we first define homogeneous type classes and functions.
There are two intuitive ideas for homogeneous type classes.
First, they have ASTs that contain only a single type variable.
Second, we can construct an FAlgebra for these type classes.

\begin{defn}
We call a type class homogeneous if:
\begin{enumerate}
\item
it has a single parameter; and
\item
the class's constraints contain only homogeneous type classes applied directly to the parameter.
\end{enumerate}
For example, all of the classes in the Prelude are homogeneous.
Some of these classes are shown in Figure \ref{code:ghc}.
Figure 2 (near the end of the paper) shows type classes which are not homogeneous due to the presence of type families in the constraints.
\end{defn}

@@

The |runSig| class method gives us a way to evaluate any |Sig|.
Mathematicians sometimes call this function simply an \emph{algebra}.
For each function constructor, |runSig| should call the corresponding function.
For each superclass constructor, |runSig| recursively calls |runSig| on the superclass's |Sig|.
This is exactly what the |Fractional| instance above does.

@@

We can also create a corresponding AST for any class that has |Num| as a superclass.
Below is the same expression encoded using |Floating|'s |FAlgebra| instance.
\begin{lstlisting}
expr2 :: AST Floating Double
expr2 = Free
  (Sig_Floating_Fractional
    (Sig_Fractional_Num
      (Sig_plus
        (Free
          (Sig_Floating_Fractional
            (Sig_Fractional_Num
              (Sig_plus
                (Pure 1)
                (Pure 2)
              )
            )
          )
        )
        (Pure 3)
      )
    )
  )
\end{lstlisting}
Notice that the |Sig_plus| constructor must be embedded into |Floating|'s |Sig| with calls to |Sig_Floating_Fractional| and |Sig_Fractional_Num|.

Constructing these ASTs by hand is a tedious process.
There is a considerable amount of boilerplate in the |Floating| AST above as a result of the need to embed the |Sig_plus| constructor.
We can avoid this boilerplate with the following type class.
\begin{lstlisting}
class
    ( FAlgebra alg1
    , FAlgebra alg2
    ) => View alg1 alg2
        where
    embedSig         :: Sig alg1 a -> Sig alg2 a
    unsafeExtractSig :: Sig alg2 a -> Sig alg1 a
\end{lstlisting}
Instances of the |View| class should satisfy the property that |alg1| is either equal to |alg2| or is a superclass of |alg2|, and
all such relationships should have a corresponding instance.
This property ensures that the |Sig alg1 a| data instance can always be embedded into a |Sig alg2 a| data instance (via a chain of superclass constructors).
It is only sometimes true that we can extract a |Sig alg1 a| from a |Sig alg2 a|,
so the function is partial and hence labeled |unsafe|.
For example,
we have the following |View Num Floating| instance because |Num| is a superclass of |Floating| via |Fractional|.
\begin{lstlisting}
instance View Num Floating
    embedSig s
        = Sig_Fractional_Floating (embedSig s)
    unsafeExtractSig (Sig_Fractional_Floating s)
        = unsafeExtractSig s
\end{lstlisting}
The |embedSig| function shows us how to embed a |Sig Num| into a |Sig Floating|;
notice that the inner call to |embedSig| relies on a |View Num Fractional| instance also existing.
The |unsafeExtractSig| function extracts the |Sig Num| instance only if the input contains one.

Armed with the |View| class, we can create a uniform representation for all class's ASTs.
The expression |(1+2)+3| can now be represented as:
\begin{lstlisting}
expr3 :: View Floating alg => AST alg Double
expr3 = Free
  (embedSig
    (Sig_plus
      (Free
        (embedSig
          (Sig_plus
            (Pure 1)
            (Pure 2)
          )
        )
      )
      (Pure 3)
    )
  )
\end{lstlisting}

Using our uniform representation of an AST,
we can make the AST itself an instance of any |FAlgebra| type class.
As an example, the |Fractional| instance is shown below.
\begin{lstlisting}
instance View Fractional alg
    => Fractional (AST alg a) where
    (/) e1 e2 = Free $ embedSig $ Sig_div e1 e2
    recip e   = Free $ embedSig $ Sig_recip e
    fromRational r
        = Free $ embedSig $ Sig_fromRational r
\end{lstlisting}
For each function in the type class,
we simply embed the function's constructor into the AST by prepending the expression |Free $ embedSig $|.

We can easily create ASTs using these instances.
The |Num| AST can be created with the expression
\begin{lstlisting}
expr4 :: AST Num Double
expr4 = (1+2)+3
\end{lstlisting}
And the |Floating| AST can be created by just changing the type signature.
\begin{lstlisting}
expr5 = AST Floating Double
expr5 = (1+2)+3
\end{lstlisting}
Notice that in these examples we didn't embed the numbers using the |Pure| constructor.
Since |AST Num Double| is an instance of |Num|,
GHC will internally call the |fromInteger| to embed the number.

\subsection{Showing the AST}
Now that we can create our ASTs, we want to actually do something with them!
The simplest task is to convert them into a |String| for display.
This is an easy procedure of writing |Show| instances for the |Sig| and |Free| data types.

@@

Now, when we type a numeric expression into ghci without a type signature,
we evaluate the expression like normal.
\begin{lstlisting}
ghci> (1+2)+3
6
\end{lstlisting}
But when we add a type signature,
we get the full AST.
\begin{lstlisting}
ghci> (1+2)+3 :: AST Floating Double
(((fromInteger 1)+(fromInteger 2))+(fromInteger 3))
\end{lstlisting}
Again, the way ghci works, all integer literals automatically have |fromInteger| applied to them,
which is why this function appears in the AST above.
It would be easy to create a pretty printer that removes these function calls and excess parentheses,
but this is left as an exercise for the reader.

\subsection{Showing functions}
Reconsider our |logLogistic1| function.
It has type
\begin{lstlisting}
logLogistic1 :: Floating a => a -> a
\end{lstlisting}
What happens if we specialize |a| to |AST Floating Double|?
We get a peek at the functions internals.
\begin{lstlisting}
ghci> logLogistic1 1 :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (fromInteger 1))))))
\end{lstlisting}
We don't just have to input single numbers;
we can also input full expressions.
These expressions will also get displayed,
intermixed with the original function.
\begin{lstlisting}
ghci> logLogistic1 (1*3+4) :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (((fromInteger 1)*(fromInteger 3))
    +(fromInteger 4)))))))
\end{lstlisting}
%\begin{lstlisting}
%ghci> logLogistic1 (logLogistic1 1) :: AST Floating Double
%(log ((fromInteger 1)/((fromInteger 1)+(exp
    %(negate (log ((fromInteger 1)/((fromInteger 1)
    %+(exp (negate (fromInteger 1)))))))))))
%\end{lstlisting}
To more conveniently display functions,
we need a variable type.
\begin{lstlisting}
newtype Var = Var String

var1 :: AST alg Var
var1 = Pure "var1"
\end{lstlisting}
Now we can show the function properly.
\begin{lstlisting}
ghci> logLogistic1 var1 :: AST Floating Var
(log ((fromInteger 1)/((fromInteger 1)+(exp (negate
    var1)))))
\end{lstlisting}
%At this point, you might be tempted to try to write a |Show| instance for functions.
%Unfortunately, you can't quite do that with this technique.

@@

To do this, we will take advantage of the free monad's structure.
The free monad is so called because as long as the parameter |f| is a |Functor|,
then |Free| has a valid |Monad| instance.
Since all |Monad|s are by necessity also |Functor|s,
|Free| has a valid |Functor| instance as well.
\begin{lstlisting}
instance Functor f => Functor (Free f) where
    fmap g (Pure a) = Pure (g a)
    fmap g (Free f) = Free (fmap (fmap g) f)
\end{lstlisting}
For our purposes, that's all the structure we'll need.

The following function evaluates the syntax tree.
\begin{lstlisting}
runAST :: (FAlgebra alg, alg a) => AST alg a -> a
runAST (Pure a) = a
runAST (Free f) = runSig (fmap evalHom f)
\end{lstlisting}
If our syntax tree already consists of just a single value (i.e. it is the |ePure| constructor),
then we just return that value.
Otherwise, we use the |Functor| instance to recursively convert the |Sig|s into types of |a|, then a call to |runSig| combines the results.

@@

ghci> runAST (logLogistic1 10 :: AST Floating Double)
-4.539889921682063e-5
\end{lstlisting}
And this demonstrates that homogeneous functions are in fact homoiconic.

@@

\section{Implementing the stabilize function}
\label{sec:stabilize}

We're now ready to modify our programs by manipulating the ASTs.
The standard way to deconstruct an AST Haskell is via pattern matching.
In this section, we'll first demonstrate how to pattern match our ASTs using the |ViewPatterns| and |PatternSynonyms| language extensions.
Then we'll see three simple examples of program transformations implemented in this way.

Automatically stabilizing floating point expressions is an active area of research.
One way of doing this is to create a library of program transformations that fix many different types of expressions.
Then combining everything in the library on your problem.
In this section, we'll create a very small library for these floating point issues,
and don't claim to have completely solved the floating point problem.
This is simply a motivating example for homoiconicity.

@@

Consider the following simple,
but numerically unstable function.
\begin{lstlisting}
testFunc1 :: Floating a => a -> a
testFunc1 = log (exp a)
\end{lstlisting}
As in our |logLogistic1| function, |testFunc| is unstable when specialized to |Double|s
because input values greater than (approximately) 710 cause the |exp| function to overflow to infinity;
the log of infinity is still infinity;
so the returned answer is infinity.
But clearly this is not the correct answer.
Mathematically, the $\log$ is the inverse of $\exp$, so the final result should be 710.
We'll write a program transformation that fixes this function,
then we'll demonstrate how the |ViewPatterns| and |PatternSynonyms| language extensions simplify this transformation.

Our goal is to convert the expression |log (exp x)| into just |x|.
To do this, we can just pattern match on the AST for the left hand side of the expression,
and replace it with the right hand side.
\begin{lstlisting}
logexpAST1 :: AST Floating a -> AST Floating a
logexpAST1 (Free (Sig_log (Free (Sig_exp a)))) = a
\end{lstlisting}
This function works, but has some warts.
First, our pattern contains noise from the need to alternate between the |Free| and |Sig_| constructors.
Second, it is specialized to the |Floating| type class,
so it won't work on any subclasses of |Floating| when it should be able to.

We'll start by fixing the second problem using the |ViewPatterns| language extension.
We will provide a very brief description of this GHC extension here;
for a more comprehensive description, see the GHC wiki \cite{viewpatterns}.
|ViewPatterns| lets us use functions within a pattern,
and we pattern match on the return value of the function.
Because functions can be polymorphic,
we can use the |ViewPatterns| extension to make our pattern accept multiple constructors.
Fortunately, we've already defined a function that extracts a |Sig| from its subclasses.
\begin{lstlisting}
unsafeExtractSig :: View alg1 alg2
    => Sig alg2 a -> Sig alg1 a
\end{lstlisting}
By using this function in the pattern, we can write a polymorphic version of our program transformation.
\begin{lstlisting}
logexpAST2 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST2
    (Free (unsafeExtractSig -> Sig_log
    (Free (unsafeExtractSig -> Sig_exp a))))
    = a
\end{lstlisting}
This function now has the correct type signature (so we've solved problem 2),
but the pattern matches are even uglier (so problem 1 is worse).
Fortunately, we can use the |PatternSynonyms| extension to simplify the patterns.
Again, see the GHC wiki for a complete description of this language extension \cite{patternsynonyms}.

We define a pattern synonym for each function contained in an FAlgebra.
Our naming convention is to prefix |AST_| to the name of the function.
These pattern synonyms combine the |Free| constructor with the view pattern from the previous example.
The synonyms for the |log| and |exp| functions are shown below.
\begin{lstlisting}
pattern AST_log :: View Floating alg
    => AST alg a -> AST alg a
pattern AST_log e
    <- Free (unsafeExtractSig -> Sig_log e)

pattern AST_exp :: View Floating alg
    => AST alg a -> AST alg a
pattern AST_exp e
    <- Free (unsafeExtractSig -> Sig_exp e)
\end{lstlisting}
Armed with these pattern synonyms,
we can rewrite our function in the much simpler form below.
\begin{lstlisting}
logexpAST3 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST3 (AST_log (AST_exp a)) = a
\end{lstlisting}
This simple function works in the simple motivating test case:
\begin{lstlisting}
ghci> logexpAST3 $ testFunc1 var1 :: AST Floating Var
var1
\end{lstlisting}
%stopzone
but it fails to work in this slightly more complex function:
\begin{lstlisting}
testFunc2 :: Floating a => a -> a
testFunc2 a = 1+exp(log a)
\end{lstlisting}
We haven't told our program transformations what to do when they don't pattern match!
The correct answer is that we want them to recurse into the subexpressions,
applying their program transformations there as well.
In the free monad, we handle these recursions using our |Functor| instance and calls to |fmap|.
\begin{lstlisting}
logexpAST4 :: View Floating alg => AST alg a -> AST alg a
logexpAST4 (AST_log (AST_exp a)) = a
logexpAST4 (Free f) = Free (fmap logexpAST4 f)
logexpAST4 (Pure a) = Pure a
\end{lstlisting}
Now we correctly recurse into subexpressions.
\begin{lstlisting}
ghci> logexpAST4 $ testFunc2 var1 :: AST Floating Var
((fromInteger 1)+var1)
\end{lstlisting}
%stopzone
There's still one last wrinkle.
What if we need to apply the program transformation multiple times?
\begin{lstlisting}
ghci> logexpAST4 $ testFunc3 var1
    :: AST Floating Var
((fromInteger 1)+(log (log (exp (exp var1)))))
\end{lstlisting}
%stopzone
We want to apply the program transformation repeatedly until it stops changing the AST.
This is called the fixed point of the transformation.
The |fixAST| function below does this.
\begin{lstlisting}
fixAST ::
    ( Eq (Sig alg (Free (Sig alg) a))
    , Eq a
    ) => (AST alg a -> AST alg a)
      -> (AST alg a -> AST alg a)
fixAST f ast = if ast==ast'
    then ast
    else fixAST f ast'
    where
        ast' = f ast
\end{lstlisting}
Now we can correctly modify the |testFunc3| function.
\begin{lstlisting}
ghci> fixAST logexpAST4 $ testFunc3 var1
    :: AST Floating Var
((fromInteger 1)+var1)
\end{lstlisting}
%stopzone

\subsection{More complex rules}

The rules in in the previous section are relatively simple.
They could just as easily have been written using GHC's rewrite rules syntax \cite{jones2001playing}.
Rewrite rules are the primary mechanism for optimising source code in GHC.
One optimization that rewrite rules cannot support is constant folding,
and there is code in GHC specially to perform this constant folding.

>>>>

\usepackage{amsmath,amsfonts}
\usepackage{url}

@@

\usepackage{courier}
%\usepackage{pxfonts}
\usepackage{listings}
\lstset{
    %language=Haskell,
    basicstyle=\ttfamily\footnotesize,
    keywordstyle=\bfseries\ttfamily\footnotesize,
    showstringspaces=false,
    %morekeywords={class,data,type,family,instance, where, ghci},
    morekeywords={ghci},
    %basicstyle=\footnotesize,
    literate={->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
             {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
    %keywordstyle=\lst@ifdisplaystyle\color{blue}\fi,
    %commentstyle=\color{gray}
}
\lstMakeShortInline|

@@

\newcommand{\homoiconic}{{\ttfamily homoiconic}~}

@@

@inproceedings{schrijvers2009complete,
  title={Complete and decidable type inference for GADTs},
  author={Schrijvers, Tom and Peyton Jones, Simon and Sulzmann, Martin and Vytiniotis, Dimitrios},
  booktitle={ACM Sigplan Notices},
  volume={44},
  number={9},
  pages={341--352},
  year={2009},
  organization={ACM}
}

@@

@article{eisenberg2015promoting,
  title={Promoting functions to type families in Haskell},
  author={Eisenberg, Richard A and Stolarek, Jan},
  journal={ACM SIGPLAN Notices},
  volume={49},
  number={12},
  pages={95--106},
  year={2015},
  publisher={ACM}
}

@@

%Imagine you've just downloaded and installed a machine learning library for Haskell.
%You open up ghci and import the corresponding modules.
%In the documentation, you see a function called |logLogistic|,
%so you ask ghci its type.
%\begin{lstlisting}
%ghci> :t logLogistic
%logLogistic :: Floating a => a -> a
%\end{lstlisting}
%Now imagine you want to know how the function is implemented.
%\begin{lstlisting}
%ghci> logLogistic var1 :: AST Floating Var
%\end{lstlisting}
%Oh no!
%The |logLogistic| function has a numerical instability in it.

Imagine you're writing a machine learning library in Haskell.
At some point, you'll need to implement the log-logistic function
\begin{equation}
f(x) = \log\left(\frac{1}{1+\exp{(-x)}}\right)
\label{eq:log-logistic}
\end{equation}
This function is used in logistic regression and many families of neural networks.
As a first attempt, you might write:
\begin{lstlisting}
logLogistic1 :: Floating x => x -> x
logLogistic1 x = log(1/(1+exp(-x)))
\end{lstlisting}
Unfortunately, this straightforward implementation is numerically unstable.
When specialized to the |Double| type,
the |exp| function overflows to infinity for inputs greater than (approximately) $709$.
%This overflow causes |logLogistic| to take the |log| of 0, which is undefined.
Therefore, on an input of $-710$,
|logLogisitic| will take the |log| of 0,
returning negative infinity when the true answer is actually very close to $-710$.
Such an inaccurate answer is unacceptable,
so the function must be rewritten.

@@

In this section we first define homogeneous type classes and functions.
There are two intuitions for homogeneous type classes.
First, they have ASTs that contain only a single type variable.
Second, we can construct an FAlgebra for these type classes.

\begin{defn}
We call a type class homogeneous if:
\begin{enumerate}
\item
it has a single parameter of kind |Type|; and
\item
the class's constraints contain only homogeneous type classes applied directly to the parameter.
\end{enumerate}
For example, all of the classes in the Prelude are homogeneous.
Some of these classes are shown in Figure \ref{code:ghc}.
Figure 2 (near the end of the paper) shows type classes which are not homogeneous due to the presence of type families in the constraints.
\end{defn}

@@

The |runSig| class method evaluates an FAlgebra's signature.
Mathematicians sometimes call this function simply an \emph{algebra}.
For each function constructor, |runSig| should call the corresponding function.
For each superclass constructor, |runSig| recursively calls |runSig| on the superclass's |Sig|.
This is exactly what the |Fractional| instance above does.

@@

We can also create an AST for the same expression using any subclass of |Num|.
Below is the same expression encoded using |Floating|'s |FAlgebra| instance.
\begin{lstlisting}
expr2 :: AST Floating Double
expr2 = Free
  (Sig_Floating_Fractional
    (Sig_Fractional_Num
      (Sig_plus
        (Free
          (Sig_Floating_Fractional
            (Sig_Fractional_Num
              (Sig_plus
                (Pure 1)
                (Pure 2)
              )
            )
          )
        )
        (Pure 3)
      )
    )
  )
\end{lstlisting}
Notice that the |Sig_plus| constructor must be embedded into |Floating|'s |Sig| with calls to |Sig_Floating_Fractional| and |Sig_Fractional_Num|.
These embeddings create considerable boilerplate,
making the construction of ASTs by hand a tedious process.

We can avoid this boilerplate with a type class that performs these embeddings for us.\footnote{
    The name {\ttfamily View} stems from the fact that the class methods are actually what GHC calls ``view patterns.''
    See Section \ref{sec:viewpatterns} for details on how the {\ttfamily ViewPatterns} language extension lets us use the {\ttfamily View} type class for easier pattern matching on our ASTs.
}
\begin{lstlisting}
class (FAlgebra alg1, FAlgebra alg2)
    => View alg1 alg2 where
    embedSig         :: Sig alg1 a -> Sig alg2 a
    unsafeExtractSig :: Sig alg2 a -> Sig alg1 a
\end{lstlisting}
Instances of |View| should satisfy the property that |alg1| is either equal to |alg2| or is a superclass of |alg2|, and
all such relationships should have a corresponding instance.
This property ensures that the |Sig alg1 a| data instance can always be embedded into a |Sig alg2 a| data instance (via a chain of superclass constructors).
It is only sometimes true that we can extract a |Sig alg1 a| from a |Sig alg2 a|,
so the function is partial and hence labeled |unsafe|.
For example,
we have the following |View Num Floating| instance because |Num| is a superclass of |Floating| via |Fractional|.
\begin{lstlisting}
instance View Num Floating
    embedSig s
        = Sig_Fractional_Floating (embedSig s)
    unsafeExtractSig (Sig_Fractional_Floating s)
        = unsafeExtractSig s
\end{lstlisting}
The |embedSig| function shows us how to embed a |Sig Num| into a |Sig Floating| by calling the appropriate superclass constructors.
Notice that the recursive call to |embedSig| relies on a |View Num Fractional| instance also existing.
Similarly, the |unsafeExtractSig| function extracts the |Sig Num| instance by pattern matching on the appropriate superclass constructors.

Armed with this |View| class, we can create a uniform representation for |FAlgebra|'s ASTs.
The expression |(1+2)+3| can now be represented polymorphically as:
\begin{lstlisting}
expr3 :: View Floating alg => AST alg Double
expr3 = Free
  (embedSig
    (Sig_plus
      (Free
        (embedSig
          (Sig_plus
            (Pure 1)
            (Pure 2)
          )
        )
      )
      (Pure 3)
    )
  )
\end{lstlisting}

Using our uniform representation of an AST,
we can make the |AST| type an instance of the homogeneous type classes.
As an example, the |Fractional| instance is shown below.
\begin{lstlisting}
instance View Fractional alg
    => Fractional (AST alg a) where
    (/) e1 e2 = Free $ embedSig $ Sig_div e1 e2
    recip e   = Free $ embedSig $ Sig_recip e
    fromRational r
        = Free $ embedSig $ Sig_fromRational r
\end{lstlisting}
For each function in the type class,
we simply embed the function's constructor into the AST by prepending the expression |Free $ embedSig $|.

We can easily create ASTs using these instances.
The |Num| AST can be created with the expression
\begin{lstlisting}
expr4 :: AST Num Double
expr4 = (1+2)+3
\end{lstlisting}
And the |Floating| AST can be created by just changing the type signature.
\begin{lstlisting}
expr5 = AST Floating Double
expr5 = (1+2)+3
\end{lstlisting}
Notice that in these last two examples we did not embed the numbers into the AST with the |Pure| constructor.
Since |AST| is an instance of |Num|,
GHC will call |fromInteger| to embed the number for us automatically.

\subsection{Showing the AST}
Now that we can create our ASTs, we want to actually do something with them!
In this section we will convert them into a |String| for display.
This is an easy procedure of writing |Show| instances for the |Sig| and |Free| data types.

@@

Now, when we type a numeric expression into ghci without a type signature,
ghci will evaluate the expression like normal.
\begin{lstlisting}
ghci> (1+2)+3
6
\end{lstlisting}
But when we add the appropriate type signature,
the full AST is displayed.
\begin{lstlisting}
ghci> (1+2)+3 :: AST Floating Double
(((fromInteger 1)+(fromInteger 2))+(fromInteger 3))
\end{lstlisting}
Again, the way ghci works, all integer literals automatically have |fromInteger| applied to them,
which is why this function appears in the AST above.
It would be easy to create a pretty printer that does not display |fromInteger| or the excess parentheses,
but this is left as an exercise for the reader.

\subsection{Showing functions}
Reconsider our |logLogistic1| function.
It has type
\begin{lstlisting}
logLogistic1 :: Floating a => a -> a
\end{lstlisting}
What happens if we specialize the |a| parameter to |AST Floating Double|?
We get a peek at the function's internals.
\begin{lstlisting}
ghci> logLogistic1 1 :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (fromInteger 1))))))
\end{lstlisting}
We don't just have to input single numbers;
we can also input full expressions.
These expressions will also get displayed,
intermixed with the original function.
\begin{lstlisting}
ghci> logLogistic1 (1*3+4) :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (((fromInteger 1)*(fromInteger 3))
    +(fromInteger 4)))))))
\end{lstlisting}
And we can even display the resulting AST of nested function calls.
\begin{lstlisting}
ghci> logLogistic1 (logLogistic1 1)
    :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (log ((fromInteger 1)/((fromInteger 1)
    +(exp (negate (fromInteger 1)))))))))))
\end{lstlisting}
It would be nice if the displayed function would display the actual input variables,
rather than just substituting numbers.
To accomplish this, we need a variable type.
\begin{lstlisting}
newtype Var = Var String
\end{lstlisting}
Values of type |Var| cannot be plugged directly into the |logLogistic1| function because |Var| is not an instance of |Floating|.
Fortunately, we have a way to make any type an instance of |Floating| via the |AST Floating| type.
More generically, we define the following variables.
\begin{lstlisting}
var1 :: AST alg Var
var1 = Pure "var1"

var2 :: AST alg Var
var2 = Pure "var2"

var3 :: AST alg Var
var3 = Pure "var3"
\end{lstlisting}
And we can use them to recover the exact AST of a function.
\begin{lstlisting}
ghci> logLogistic1 var1 :: AST Floating Var
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate var1)))))
\end{lstlisting}
At this point, you might be tempted to write a |Show| instance for functions.
\begin{lstlisting}
instance
    ( FAlgebra alg
    , Show (Sig alg (Free (Sig alg) Var))
    ) => Show (AST alg Var -> AST alg Var) where
    show f = show (f var1)
\end{lstlisting}
Unfortunately, if you try to use the instance naively
\begin{lstlisting}
ghci> logLogistic1
\end{lstlisting}
then ghci will complain about ambiguous type variables.
GHC's defaulting mechanism is not powerful enough for it to properly specialize the function without a type signature.
So unfortunately this technique doesn't give us a convenient |Show| instance,
it does give us a slightly more verbose alternative for showing a function.

@@

To do this, we will take advantage of the free monad's structure.
(Recall that the |AST| type we've been using is a synonym for the free monad.)
The free monad is so called because as long as the parameter |f| is a |Functor|,
then |Free| has a valid |Monad| instance.
Since all |Monad|s are by necessity also |Functor|s,
|Free| has a valid |Functor| instance as well.
\begin{lstlisting}
instance Functor f => Functor (Free f) where
    fmap g (Pure a) = Pure (g a)
    fmap g (Free f) = Free (fmap (fmap g) f)
\end{lstlisting}
For our purposes, that's all the structure we'll need.

The following simple function evaluates the syntax tree.
\begin{lstlisting}
runAST :: (FAlgebra alg, alg a) => AST alg a -> a
runAST (Pure a) = a
runAST (Free f) = runSig (fmap evalHom f)
\end{lstlisting}
If our syntax tree already consists of just a single value (i.e. it is the |Pure| constructor),
then we just return that value.
Otherwise, we use the |Functor| instance to recursively convert the |Sig|s into types of |a|,
combining the results with |runSig|.

@@

ghci> runAST (logLogistic1 10 :: AST Floating Double)
-4.539889921682063e-5
\end{lstlisting}
We've finally demonstrated that homogeneous functions are in fact homoiconic.

@@

\section{Writing program transformations}
\label{sec:stabilize}

We're now ready to write program transformations that manipulate the ASTs.
The standard way to deconstruct an AST Haskell is via pattern matching.
In this section, we'll first demonstrate how to pattern match our ASTs using the |ViewPatterns| and |PatternSynonyms| language extensions.
Then we'll see three simple examples of program transformations implemented in this way.

%Automatically stabilizing floating point expressions is an active area of research.
%One way of doing this is to create a library of program transformations that fix many different types of expressions.
%Then combining everything in the library on your problem.
%In this section, we'll create a very small library for these floating point issues,
%and don't claim to have completely solved the floating point problem.
%This is simply a motivating example for homoiconicity.

@@

Consider the following simple,
but numerically unstable function.
\begin{lstlisting}
testFunc1 :: Floating a => a -> a
testFunc1 = log (exp a)
\end{lstlisting}
As in our |logLogistic1| function, |testFunc| gives incorrect answers when specialized to |Double|.
Inputs greater than (approximately) 710 cause the |exp| function to overflow to infinity;
the log of infinity is still infinity;
so the returned answer is infinity.
But this wrong.
Mathematically, the $\log$ function is defined to be the inverse of $\exp$,
so the final result should be exactly the same as the input.
We'll write a program transformation that stabilizes the |testFunc1| function,
then we'll demonstrate how the |ViewPatterns| and |PatternSynonyms| language extensions simplify this transformation.
\label{sec:viewpatterns}

Our goal is to take in input AST,
find all occurrences of the expression |log (exp x)|,
and replace them with just |x|.
Here |x| can be an arbitrary subtree of the original AST.
We will create a series of four functions to accomplish this task,
each improving on the previous.
Here is the first.
\begin{lstlisting}
logexpAST1 :: AST Floating a -> AST Floating a
logexpAST1 (Free (Sig_log (Free (Sig_exp a)))) = a
\end{lstlisting}
This function has two syntactic warts.
First, the pattern is noisy.
To properly match the expression, we must alternate between the |Free| and |Sig_| constructors.
Second, the function is specialized to the |Floating| type class.
We would like our transformation to also works on subclasses of |Floating|.

We'll start by correcting the second problem using the |ViewPatterns| language extension.
|ViewPatterns| provides a syntactic sugar that lets us pattern match on the return value of a function.\footnote{
    The GHC wiki \cite{viewpatterns} provides details on exactly how the desugaring works.
}
We can use |ViewPatterns| to write polymorphic patterns by using polymorphic functions in the pattern.
Fortunately, we've already defined the exact function we need:
\begin{lstlisting}
unsafeExtractSig :: View alg1 alg2
    => Sig alg2 a -> Sig alg1 a
\end{lstlisting}
By using this function in the pattern,
we can write a polymorphic version of our program transformation.
\begin{lstlisting}
logexpAST2 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST2
    (Free (unsafeExtractSig -> Sig_log
    (Free (unsafeExtractSig -> Sig_exp a))))
    = a
\end{lstlisting}
This function now has the correct type signature (so we've solved problem 2),
but the pattern matches are even uglier (so problem 1 is worse).

The |PatternSynonyms| extension lets us define new patterns,
and we will use this feature to clean up our syntax.\footnote{
    Again, see the GHC Wiki \cite{patternsynonyms} for details on this language extension.
}
Specifically, we define a pattern synonym for each function contained in an FAlgebra.
Our naming convention is to prefix |AST_| to the name of the function.
These pattern synonyms combine the |Free| constructor with the view pattern from the previous example.
The synonyms for the |log| and |exp| functions are shown below.
\begin{lstlisting}
pattern AST_log :: View Floating alg
    => AST alg a -> AST alg a
pattern AST_log e
    <- Free (unsafeExtractSig -> Sig_log e)

pattern AST_exp :: View Floating alg
    => AST alg a -> AST alg a
pattern AST_exp e
    <- Free (unsafeExtractSig -> Sig_exp e)
\end{lstlisting}
Armed with these pattern synonyms,
we can rewrite our function in a much simpler form.
\begin{lstlisting}
logexpAST3 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST3 (AST_log (AST_exp a)) = a
\end{lstlisting}
This simple function works on our the motivating test case:
\begin{lstlisting}
ghci> logexpAST3 $ testFunc1 var1 :: AST Floating Var
var1
\end{lstlisting}
%stopzone
but it fails to work in this slightly more complex function:
\begin{lstlisting}
testFunc2 :: Floating a => a -> a
testFunc2 a = 1+exp(log a)
\end{lstlisting}
We haven't told our program transformations what to do when they don't pattern match!
The correct answer is that we want the transformations to recurse into the subexpressions,
applying the transformations there as well.
In the free monad, recursion is handled by a call to |fmap| and the base case is handled by extracting the value of |Pure|.
This gives our final version of the program transformation.
\begin{lstlisting}
logexpAST4 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST4 (AST_log (AST_exp a)) = a
logexpAST4 (Free f) = Free (fmap logexpAST4 f)
logexpAST4 (Pure a) = Pure a
\end{lstlisting}
Now we correctly recurse into subexpressions.
\begin{lstlisting}
ghci> logexpAST4 $ testFunc2 var1
    :: AST Floating Var
((fromInteger 1)+var1)
\end{lstlisting}
%stopzone
There's still one last wrinkle.
What if we need to apply the program transformation multiple times?
Consider the following test function.
\begin{lstlisting}
testFunc3 :: Floating a => a -> a
testFunc3 a = 1+log(log(log(exp(exp(exp a)))))
\end{lstlisting}
Applying our program transformation directly doesn't fix the function because the transformation needs to be applied repeatedly.
\begin{lstlisting}
ghci> logexpAST4 $ testFunc3 var1
    :: AST Floating Var
((fromInteger 1)+(log (log (exp (exp var1)))))
\end{lstlisting}
%stopzone
What we want is the \emph{fixed point} of the program transformation.
That is, we want to apply the program transformation repeatedly until it stops changing the AST.
The |fixAST| function below takes as input a transformation and returns the fixed point of that transformation.
\begin{lstlisting}
fixAST ::
    ( Eq (Sig alg (Free (Sig alg) a))
    , Eq a
    ) => (AST alg a -> AST alg a)
      -> (AST alg a -> AST alg a)
fixAST f ast = if ast==ast'
    then ast
    else fixAST f ast'
    where
        ast' = f ast
\end{lstlisting}
Now we can correctly modify the |testFunc3| function.
\begin{lstlisting}
ghci> fixAST logexpAST4 $ testFunc3 var1
    :: AST Floating Var
((fromInteger 1)+var1)
\end{lstlisting}
%stopzone

\subsection{More complex transformations}
